---
title: "Exploratory Analysis — Peripheral Gaze-Change Detection"
author: "Amira Shymbolatova"
date: last-modified
format: html
execute:
  echo: true
  warning: false
  message: false
---

# Setup

```{r}
library(tidyverse)

# ---- SET THIS TO YOUR TRIALS CSV FILE PATH ----
TRIALS_FILE <- "trials_GazePeriphery_XXXXXXXX.csv"
PARTICIPANT_ID <- "pilot_1"   # change per participant
```

# Data reading

```{r}
# Read and prepare data
# Adapated from Schmettow (2026) GR_1.qmd
# Key differences from supervisor's version:
#   - your data has 'condition' (control/quadbright) instead of blockNumber/eccentricity
#   - your data has 'magnitude' (gaze-change size) instead of 'angle'
#   - your data has 'eccentricity' as continuous variable (5-28 degrees)
#   - filter valid == 1 to exclude fixation breaks

D_0 <- read_csv(TRIALS_FILE, show_col_types = FALSE)

D_1 <- D_0 |>
  filter(valid == 1) |>
  mutate(
    Part        = as.factor(PARTICIPANT_ID),
    condition   = as.factor(condition),
    eccentricity = as.numeric(eccentricity),
    magnitude   = as.numeric(magnitude),
    correct     = as.numeric(correct)
  ) |>
  group_by(condition) |>
  mutate(trial = row_number()) |>
  ungroup() |>
  select(Part, trial, condition, eccentricity, magnitude, correct,
         srt_ms, keyboard_rt_ms, pre_change_ms)

summary(D_1)
```

# Exploration

## Plot 1 — Proportion correct vs eccentricity per condition

This is the key plot your supervisor requested. It shows the sigmoid curve 
approaching 0.5 (chance) at high eccentricities.

```{r}
#| fig-width: 8
#| fig-height: 5

# Summarize proportion correct per eccentricity per condition
# (equivalent to supervisor's T_1 but split by condition)
T_1 <- D_1 |>
  summarize(
    p_correct = mean(correct),
    n_Obs     = n(),
    .by       = c(condition, eccentricity)
  )

T_1 |>
  ggplot(aes(x = eccentricity, 
             color = condition, 
             y = p_correct)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", 
             color = "gray50") +
  geom_point(aes(size = n_Obs), alpha = 0.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.8) +
  scale_y_continuous(limits = c(0.3, 1.05),
                     labels = scales::percent_format()) +
  scale_color_manual(values = c("control" = "#2196F3", 
                                "quadbright" = "#FF5722")) +
  labs(
    title    = paste("Proportion correct by eccentricity —", PARTICIPANT_ID),
    subtitle = "Dashed line = chance (50%)",
    x        = "Eccentricity (degrees from fixation)",
    y        = "Proportion correct",
    color    = "Condition",
    size     = "N trials"
  ) +
  theme_minimal(base_size = 13)
```

## Plot 2 — Proportion correct vs magnitude per condition

Equivalent to supervisor's main plot. Shows how detection improves 
with larger gaze-change magnitude.

```{r}
#| fig-width: 8
#| fig-height: 5

# Bin magnitude like supervisor binned angle (0-10, 10-15, 15-20)
T_2 <- D_1 |>
  mutate(Magnitude_range = case_when(
    magnitude < 5  ~ "1-4",
    magnitude < 10 ~ "5-9",
    magnitude < 15 ~ "10-14",
    magnitude < 20 ~ "15-19",
    magnitude < 25 ~ "20-24",
    TRUE           ~ "25-30"
  )) |>
  summarize(
    p_correct = mean(correct),
    n_Obs     = n(),
    .by       = c(condition, Magnitude_range, eccentricity)
  )

T_2 |>
  ggplot(aes(x = Magnitude_range, 
             color = as.factor(round(eccentricity)), 
             y = p_correct)) +
  facet_wrap(~condition) +
  geom_point(aes(size = n_Obs), alpha = 0.7) +
  geom_line(aes(group = round(eccentricity)), alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_y_continuous(limits = c(0.3, 1.05),
                     labels = scales::percent_format()) +
  labs(
    title  = paste("Proportion correct by magnitude range —", PARTICIPANT_ID),
    x      = "Gaze-change magnitude (degrees)",
    y      = "Proportion correct",
    color  = "Eccentricity (°)",
    size   = "N trials"
  ) +
  theme_minimal(base_size = 12)
```

## Plot 3 — Eccentricity curve per condition (participant level)

Direct equivalent of supervisor's per-participant plot.
Used for participant rejection criterion.

```{r}
#| fig-width: 8
#| fig-height: 5

D_1 |>
  mutate(Magnitude_range = case_when(
    magnitude < 10 ~ "small (1-9)",
    magnitude < 20 ~ "medium (10-19)",
    TRUE           ~ "large (20-30)"
  )) |>
  summarize(
    p_correct = mean(correct),
    n_Obs     = n(),
    .by       = c(Part, condition, Magnitude_range, eccentricity)
  ) |>
  ggplot(aes(x = eccentricity, 
             color = condition,
             y = p_correct)) +
  facet_wrap(~Magnitude_range) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  geom_point(aes(size = n_Obs), alpha = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 1.0) +
  scale_y_continuous(limits = c(0.2, 1.05),
                     labels = scales::percent_format()) +
  scale_color_manual(values = c("control" = "#2196F3",
                                "quadbright" = "#FF5722")) +
  labs(
    title    = paste("Detection by eccentricity and magnitude —", PARTICIPANT_ID),
    subtitle = "Participant rejection: curve should be monotonically decreasing toward 50%",
    x        = "Eccentricity (degrees from fixation)",
    y        = "Proportion correct",
    color    = "Condition",
    size     = "N trials"
  ) +
  theme_minimal(base_size = 12)
```

## Participant rejection check

Based on supervisor's criterion:
> "Legit participants should show performance above chance level 
> and have a clear upwards trend, as well as the expected ordering 
> of eccentricity."

```{r}
# Check if participant shows above-chance performance overall
overall_acc <- D_1 |>
  group_by(condition) |>
  summarize(
    mean_correct = mean(correct),
    n_trials     = n(),
    above_chance = mean(correct) > 0.55
  )

cat("Overall accuracy by condition:\n")
print(overall_acc)

# Check monotonicity: does accuracy decrease with eccentricity?
monotonicity <- D_1 |>
  group_by(condition, eccentricity) |>
  summarize(p_correct = mean(correct), .groups = "drop") |>
  group_by(condition) |>
  summarize(
    cor_with_ecc = cor(eccentricity, p_correct, method = "spearman"),
    monotone_decreasing = cor(eccentricity, p_correct, method = "spearman") < -0.3
  )

cat("\nMonotonicity check (should be negative correlation):\n")
print(monotonicity)

# Overall rejection decision
reject <- any(!overall_acc$above_chance) | any(!monotonicity$monotone_decreasing)
cat("\nParticipant rejection flag:", ifelse(reject, "CONSIDER REJECTION", "KEEP"), "\n")
```

# Statistical model

Adapted directly from supervisor's simplified psychophysics model (GR_1.qmd).

Key adaptation: added `condition` as predictor for dprime, since your study
compares control vs quadbright conditions. This is your main hypothesis.

```{r}
#| eval: false

# Install bayr if needed:
# devtools::install_github("schmettow/bayr")

library(brms)
library(bayr)

# Inverse logit helper (from supervisor's code)
inv_logit <- function(x) exp(x) / (1 + exp(x))

# Psychometric function (from supervisor's code)
# guess = 0.5 for 2AFC
# dprime models sensitivity as function of magnitude AND eccentricity AND condition
# lapse and bias dropped (simplified model, as supervisor did)

F_1 <- bf(
  correct ~ 0.5 + (1 - 0.5) * inv_logit(exp(dprime) * magnitude),
  dprime ~ 1 + eccentricity + condition + pre_change_ms +
           (1 + eccentricity | Part),
  nl = TRUE
)

# Priors (from supervisor's code)
Priors_1 <- c(
  set_prior("normal(0, 1)", nlpar = "dprime")
)

# Fit model
M_1 <- brm(
  F_1,
  data   = D_1,
  family = bernoulli(link = "identity"),
  prior  = Priors_1,
  chains = 4,
  cores  = 4,
  iter   = 3000,
  warmup = 2000,
  control = list(adapt_delta = 0.99, max_treedepth = 12)
)

save(M_1, D_1, file = "M_1_pilot.Rda")
```

```{r}
#| eval: false

# After fitting — extract and plot posteriors
load("M_1_pilot.Rda")
P_1 <- bayr::posterior(M_1)

# Population-level estimates
fixef(P_1)

# Posterior distributions
P_1 |>
  filter(type == "fixef") |>
  ggplot(aes(x = value, col = parameter)) +
  facet_grid(~nonlin) +
  geom_density() +
  theme_minimal()
```
